---
title: "Atom plots"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{atom-plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
require(tlf)
```

```{r load theme, echo=FALSE}
vignetteTheme <- loadThemeFromJson(system.file("extdata", "template-theme.json", package = "tlf"))
useTheme(vignetteTheme)
```
This vignette tackles how to use atom plots to perform simple to more complex plots with the `tlf`-library.

# Introduction to atom plots

Atom plots correspond to simple plots that can be used independently or in combination to form a final plot, often more complex, that will be referred as molecule plot.
In the sequel, the following atoms are detailed:

- *initializePlot*
- *addScatter*
- *addLine*
- *addRibbon*
- *addErrorbar*

# Initialize a plot object: *initializePlot*

The `tlf` method *initializePlot* aims at creating an empty plot with properties defined by a *PlotConfiguration* R6 class or subclass object.
More details on *PlotConfiguration* objects are available in a separate vignette.
The default option for *initializePlot()* does not require a *plotConfiguration* object to be specifically input by the user; 
in which case, *plotConfiguration* will be derived automatically from the current theme.

The output plot from *initializePlot()* is a regular `ggplot` object to which a *plotConfiguration* field is inclduded.
Since this method is at the core of every atom and molecule plots of the `tlf`-library, any plot generated using this package will include a *plotConfiguration* field.

Below are a few examples of use case of *initializePlot*:
```{r initializePlot}
# Inintialize an empty plot without specific configuration, use the current theme
emptyPlot <- initializePlot()

# Create a PlotConfiguration object with some specifications
myConfiguration <- PlotConfiguration$new(title = "My empty plot",
                                         xlabel = "X",
                                         ylabel = "Y")
# Use the specification to create the plot
emptyPlotWithLabels <- initializePlot(myConfiguration)

# PlotConfiguration class has a smart mapping feature,
# directly deriving labels from data and its metaData and its dataMapping
time <- seq(0,20, 0.1)
myData <- data.frame(x = time,
                     y = 2*cos(time))
# If metaData includes dimension and unit, 
# they can be use by the smart configuration to generate labels as "dimension [unit]"
myMetaData <- list(x = list(dimension = "Time",
                            unit = "min"),
                   y = list(dimension = "Amplitude",
                            unit = "cm"))
myMapping <- XYGDataMapping$new(x = "x",
                                y = "y")

# In this example, the configuration will use the x/y mapping from dataMapping
# to fetch dimension and unit from metaData
mySmartConfiguration <- PlotConfiguration$new(title = "My smart plot",
                                              data = myData,
                                              metaData = myMetaData,
                                              dataMapping = myMapping)

smartEmptyPlot <- initializePlot(mySmartConfiguration)
```

```{r initializePlot output, echo=FALSE, fig.cap="Left: emptyPlot; Middle: emptyPlotWithLabels; Right: smartEmptyPlot", fig.width=7.5}
gridExtra::grid.arrange(emptyPlot, emptyPlotWithLabels, smartEmptyPlot, ncol=3)
```

# Add scatter or lines to a plot object: *addScatter*, *addLine*

The functions *addScatter* and *addLine* aims at providing an easy interface to add points or lines to a plot object.
If a plot object is specified in *plotObject* input, then the layer will be added on top of the previous plot.
In the case the input is `NULL` (default value), the function *initializePlot* is called internally.

The following data will be used as an example in the next section:
```{r get ExampleData, echo=FALSE}
myTestData <- read.csv(system.file("extdata", "test-data.csv", package = "tlf"), stringsAsFactors = FALSE)
myTestData <- data.frame(x = myTestData$Obs, 
                         y = myTestData$Pred,
                         type = myTestData$Sex,
                         country = myTestData$Country) 
```
```{r show ExampleData, results='asis'}
knitr::kable(utils::head(myTestData)) 
```

## *addScatter*

The primary input parameter of function *addScatter* is *data* and can be used along with its *metaData* and *dataMapping*.
However, it is also possible to directly use the input parameters *x* and *y* instead to plot directly *y* as a function of *x*.

```{r scatter mapping}
# Naive mapping on x and y
myMapping <- XYGDataMapping$new(x = "x",
                                y = "y")

# Mapping on x and y, while grouping the data by type
myGroupMapping <- XYGDataMapping$new(x = "x",
                                     y = "y",
                                     color = "type")
# Using the naive mapping
pNaive <- addScatter(data = myTestData,
                     metaData = myMetaData,
                     dataMapping = myMapping)

# Using tlf smart mapping (same as naive)
pSmart <- addScatter(data = myTestData,
                     metaData = myMetaData)

# Using group in mapping
pGroup <- addScatter(data = myTestData,
                metaData = myMetaData,
                dataMapping = myGroupMapping)

# Using x and y (same as naive but without metaData, no label)
pXY <- addScatter(x = myTestData$x,
                  y = myTestData$y)
```

```{r show scatter mapping, echo=FALSE, fig.cap="top left: pNaive; top right: pSmart; bottom left: pGroup; bottom right: pXY", fig.width=7.5}
gridExtra::grid.arrange(pNaive, pSmart, pGroup, pXY, ncol=2)
```

Input parameters of *addScatter* also include *plotConfiguration* and *plotObject* as previously described.
Input parameters *caption*, *color*, *size*, *shape* and *linetype* can overwrite the default aesthetic and legend values defined by *plotConfiguration*.
These parameters are optional and will be `NULL` by default, which will lead to use aesthetic from the current theme.
In particular, *dataMapping* also allows to create a legend caption, the input parameter *caption* can overwrtie such legend caption.

```{r addScatter options}
males <- myTestData$type %in% "Male"
females <- myTestData$type %in% "Female"
# Using x and y
pScatter1 <- addScatter(x = myTestData$x[males],
                        y = myTestData$y[males],
                        color = "blue",
                        size = 3,
                        caption = "Male")
pScatter2 <- addScatter(x = myTestData$x[females],
                        y = myTestData$y[females],
                        color = "pink", 
                        size = 3,
                        caption = "Female")
pScatter3 <- addScatter(x = myTestData$x[females],
                        y = myTestData$y[females],
                        color = "pink", 
                        size = 3,
                        caption = "Female",
                        plotObject = pScatter1)
```

```{r, echo=FALSE, fig.cap="Left: pScatter1; Middle: pScatter2; Right: pScatter3", fig.width=7.5}
gridExtra::grid.arrange(pScatter1, pScatter2, pScatter3, ncol=3)
```

Finally, the last input parameters include *caption*, *color*, *size*, *shape* and *linetype*. 
These parameters are optional and will use the default of the current theme if they are not defined.
In particular, *dataMapping* also allows to create a legend caption, the input parameter *caption* can overwrtie such legend caption.

## *addLine*

The function *addLine* is almost identical to *addScatter*.
It differs regarding its default aesthetic properties: lines are replacing the points, even if points can be used.
It also differs by the possibility of using *x* or *y* input for x/y-intercepts.

Using a different example,

```{r addLine options}
# Using x and y
pLine1 <- addLine(x = seq(1,5,0.1),
                  y = cos(seq(1,5,0.1)),
                  color = "blue",
                  linetype = "solid",
                  caption = "cosinus")
pLine2 <- addLine(y = c(-1, 1),
                  color = "pink",
                  linetype = "longdash",
                  caption = "extremes")
pLine3 <- addLine(y = c(-1, 1),
                  color = "pink",
                  linetype = "longdash",
                  caption = "extremes",
                  plotObject = pLine1)
```

```{r show addLine, echo=FALSE, fig.cap="Left: pLine1; Middle: pLine2; Right: pLine3", fig.width=7.5}
gridExtra::grid.arrange(pLine1, pLine2, pLine3, ncol=3)
```

# Add ribbon or errorbars to a plot object: *addRibbon*, *addErrorbar*

These 2 atoms work in a similar fashion to *addScatter* and *addLine* but they use a different mapping that require *ymin* and *ymax* insteada of *y*.

## *addRibbon*

## *addErrorbar*

