---
title: "Atom plots"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{atom-plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tlf)
```

This vignette tackles how to use atom plots to perform simple to more complex plots with the `tlf`-library.

# Introduction to atom plots

Atom plots correspond to simple plots that can be used independently or in combination to form a final plot, often more complex, that will be referred as molecule plot.
In the sequel, the following atoms are detailed:

- *initializePlot*
- *addScatter*
- *addLine*
- *addRibbon*
- *addErrorbar*

# Initialize a plot object: *initializePlot*

The `tlf` method *initializePlot* aims at creating an empty plot with properties defined by a  *PlotConfiguration* R6 class or subclass object.
The output plot is a regular `ggplot` object to which a *plotConfiguration* field is implanted.
Since this method is at the core of every atom and molecule plots, any plot generated using the `tlf`-library will include such a *plotConfiguration*.

If no *plotConfiguration* is provided to the method *initializePlot*, the default configuration from the current theme will be used.

Below are a few examples of use case of *initializePlot*:
```{r}
# No specific configuration
emptyPlot <- initializePlot()

# PlotConfiguration
myConfiguration <- PlotConfiguration$new(title = "My empty plot",
                                         xlabel = "X",
                                         ylabel = "Y")
emptyPlotWithLabels <- initializePlot(myConfiguration)

# Smart PlotConfiguration
time <- seq(0,20, 0.1)
myData <- data.frame(x = time,
                     y = 2*cos(time))
myMetaData <- list(x = list(dimension = "Time",
                            unit = "min"),
                   y = list(dimension = "Amplitude",
                            unit = "cm"))
myMapping <- XYGDataMapping$new(x = "x",
                                y = "y")

mySmartConfiguration <- PlotConfiguration$new(title = "My smart plot",
                                              data = myData,
                                              metaData = myMetaData,
                                              dataMapping = myMapping)

smartEmptyPlot <- initializePlot(mySmartConfiguration)

# Using another theme to change default
useTheme(tlfTheme)
myNewSmartConfiguration <- PlotConfiguration$new(title = "My smart plot",
                                              data = myData,
                                              metaData = myMetaData,
                                              dataMapping = myMapping)
smartEmptyPlotNewTheme <- initializePlot(myNewSmartConfiguration)
```

```{r, echo=FALSE}
useTheme(defaultTheme)
```

```{r, echo=FALSE, fig.cap="top left: emptyPlot; top right: emptyPlotWithLabels; bottom left: smartEmptyPlot; bottom right: ", fig.width=7.5}
gridExtra::grid.arrange(emptyPlot, emptyPlotWithLabels, smartEmptyPlot, smartEmptyPlotNewTheme, ncol=2)
```

# Add scatter or lines to a plot object: *addScatter*, *addLine*

The functions *addScatter* and *addLines* can be used to add scatter and line layers.
They can be used directly, in which case the function *initializePlot* will be called from within these functions. Or they can be added on top of a previous plot object using the parameter *plotObject*.

The following data will be used as an example in the next section:
```{r}
myCosData <- rbind.data.frame(data.frame(x = time,
                                         y = 2*cos(time),
                                         type = "cosinus"),
                              data.frame(x = time,
                                         y = 2*sin(time),
                                         type = "sinus"),
                              data.frame(x = time,
                                         y = round(2*sin(time)),
                                         type = "stairs"))
```

## *addScatter*

The primary input parameter of function *addScatter* is *data* and can be used along with its *metaData* and *dataMapping*.
However, it is also possible to directly use the input parameters *x* and *y* instead to plot directly *y* as a function of *x*.

```{r}
# Naive mapping on x and y
myMapping <- XYGDataMapping$new(x = "x",
                                y = "y")

# Mapping on x and y, while grouping the data by type
myGroupMapping <- XYGDataMapping$new(x = "x",
                                     y = "y",
                                     color = "type")
# Using the naive mapping
pNaive <- addScatter(data = myCosData,
                     metaData = myMetaData,
                     dataMapping = myMapping)

# Using tlf smart mapping (same as naive)
pSmart <- addScatter(data = myCosData,
                     metaData = myMetaData)

# Using group in mapping
pGroup <- addScatter(data = myCosData,
                metaData = myMetaData,
                dataMapping = myGroupMapping)

# Using x and y (same as naive but without metaData, no label)
pXY <- addScatter(x = myCosData$x,
                  y = myCosData$y)
```

```{r, echo=FALSE, fig.cap="top left: pNaive; top right: pSmart; bottom left: pGroup; bottom right: pXY", fig.width=7.5}
gridExtra::grid.arrange(pNaive, pSmart, pGroup, pXY, ncol=2)
```

In the other input parameters, *plotConfiguration* or a previous *plotObject* can be used to define the either which will be the plot configuration or on top of which previous plot the scatter layer should be added.

```{r}
# Using x and y
pScatter1 <- addScatter(x = myCosData$x[myCosData$type %in% "cosinus"],
                  y = myCosData$y[myCosData$type %in% "cosinus"])


```

```{r, echo=FALSE, fig.cap="top left: pNaive; top right: pSmart; bottom left: pGroup; bottom right: pXY", fig.width=7.5}
gridExtra::grid.arrange(pScatter1, pScatter1, pScatter1, pScatter1, ncol=2)
```

Finally, the last input parameters include *caption*, *color*, *size*, *shape* and *linetype*. 
These parameters are optional and will use the default of the current theme if they are not defined.
In particular, *dataMapping* also allows to create a legend caption, the input parameter *caption* can overwrtie such legend caption.

## *addLine*

The function *addLine* is almost identical to *addScatter*.
It differs regarding its default aesthetic properties: lines are replacing the points, even if points can be used.
It also differs by the possibility of using only *y* input for y-intercepts.

Using the example with *myData*,

```{r}

line1 <- addLine(data = myCosData[myCosData$type %in% "cosinus",],
                 metaData = myMetaData,
                 dataMapping = myMapping,
                 caption = "cosinus")
# y intercepts
line2 <- addLine(y = c(-2,0,2),
                 caption = "y-intercepts",
                 color = "darkblue",
                 size = 2,
                 linetype = "longdash",
                 plotObject = line1)

line3 <- addLine(data = myCosData,
                 metaData = myMetaData,
                 dataMapping = myGroupMapping)

line4 <- addLine(x = seq(0, 5*pi, pi),
                 y = rep(c(-1, 1), 3),
                 caption = "frame",
                 color = "brown4",
                 linetype = "solid",
                 plotObject = line3)

```

```{r, echo=FALSE, fig.cap="top left: line1; top right: line2; bottom left: line3; bottom right: line4", fig.width=7.5}
gridExtra::grid.arrange(line1, line2, line3, line4, ncol=2)
```

# Add ribbon or errorbarsa plot object: *addRibbon*, *addErrorbar*

