---
title: "box-whisker-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{box-whisker-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tlf)
```


# 1. Introduction
## 1.1. Objectives
The following vignette aims at documenting and illustrating workflows for producing PK ratio plots using the `tlf`-Library.

## 1.2. Prerequisite
This vignette focuses PK ratio plots examples. Detailed documentation on typical `tlf` workflow, use of `AgregationSummary`, `DataMapping`, `PlotConfiguration` and `Theme` can be found in `vignette("tlf-workflow")`.

# 2. Illustration of basic Boxplots
## 2.1. Data
```{r}
boxData <- data.frame(
  x = c(rep("A", 100), rep("B", 100), rep("C", 100)),
  y = c(rnorm(100) + 5, 10 + rnorm(100), 10 - 2 * rnorm(100))
)
```
```{r, echo=FALSE}
knitr::kable(head(boxData))
```

## 2.2. `plotBoxWhisker`

## 2.3. Minimal example
```{r, fig.height=5, fig.width=7.5}
boxMap <- BoxWhiskerDataMapping$new(x = "x",
                                    y = "y")

boxplot <- plotBoxWhisker(boxData, 
                          dataMapping = boxMap)
boxplot
```

## 2.4. Tune Boxes, Whiskers and Outliers functions
```{r, fig.height=5, fig.width=7.5}
normMap <- BoxWhiskerDataMapping$new(x = "x",
                                    y = "y",
                                    ymin = tlfStatFunctions$`mean-1.96sd`,
                                    lower = tlfStatFunctions$`mean-sd`,
                                    middle = tlfStatFunctions$mean,
                                    upper = tlfStatFunctions$`mean+sd`,
                                    ymax = tlfStatFunctions$`mean+1.96sd`)

normplot <- plotBoxWhisker(boxData, 
                          dataMapping = normMap)
normplot
```

```{r, fig.height=5, fig.width=7.5}
outMap <- BoxWhiskerDataMapping$new(x = "x",
                                    y = "y",
                                    minOutlierLimit = tlfStatFunctions$`Percentile-2.5%`,
                                    maxOutlierLimit = tlfStatFunctions$`Percentile-97.5%`)

outplot <- plotBoxWhisker(boxData, 
                          dataMapping = outMap)
outplot
```
