---
title: "Box-whisker plots"
author: "OSPSuiteR 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{box-whisker-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup}
require(tlf)
```

```{r load theme, echo=FALSE}
vignetteTheme <- loadThemeFromJson(system.file("extdata", "template-theme.json", package = "tlf"))
useTheme(vignetteTheme)
```

# 1. Introduction
## 1.1. Objectives
The following vignette aims at documenting and illustrating workflows for producing PK ratio plots using the `tlf`-Library.

## 1.2. Prerequisite
This vignette focuses boxplot examples. Detailed documentation on typical `tlf` workflow, use of `AgregationSummary`, `DataMapping`, `PlotConfiguration` and `Theme` can be found in `vignette("tlf-workflow")`.

# 2. Definition of the boxplot fonctions and classes
## 2.1. The `plotBoxWhisker` function
The function for plotting box-whiskers is: `plotBoxWhisker`. 
Basic documentation of the function can be found using: `?plotBoxWhisker`.
The typical usage of this function is: `plotBoxWhisker(data, metaData = NULL, dataMapping = NULL, plotConfiguration = NULL)`.
The output of the function is a `ggplot` object.

## 2.2. The `BoxWhiskerDataMapping` class
The `dataMapping` from `plotBoxWhisker` requires a `BoxWhiskerDataMapping` class.
This class can simply be initialized by `BoxWhiskerDataMapping$new()`, needing `y` variable name input only.
For boxplots with multiple boxes, `x` variable name or/and `fill` groupMapping can be used. The `x` variable is expected to be factor levels.
Beside these common input, it is possible to overwrite the aggregation functions that plot the edges of the box, the whiskers and the oultying data.

- For the box edges `lower`, `middle` and `upper` use respectively the first quartile, median and third quartile (25th, 50th and 75th percentiles).
- For the whiskers, `ymin` and `ymax` use the 5th and 95th percentiles.
- For outliers, points lower than the 25th percentile - 1.5IQR and points higher than 75th percentiles + 1.5IQR (where IQR is the inter-quartile range) are flagged and plotted.

In order to help with the boxplot aggregation functions, a bank of predefined function names is already available in the tlfStatFunctions (as an enum). Consequently, a tree with the available predefined function names will appear when writing `tlfStatFunctions$`: `r paste0("'", names(tlfStatFunctions), "'", sep = ", ", collapse = "")`

# 3. Examples
## 3.1. Data
To illustrate the workflow leading to performing boxplots. The `pkRatioDataExample.RData` example can be loaded from the `data` folder. 
It includes the dataset `pkRatioData` and its metaData `pkRatioMetaData`.

```{r load data, results='asis'}
# Load example
pkRatioData <- read.csv(system.file("extdata", "test-data.csv", package = "tlf"), stringsAsFactors = FALSE)

# pkRatioData
knitr::kable(utils::head(pkRatioData), digits = 2)
```

```{r load metadata, echo=FALSE}
# Load example
pkRatioMetaData <- list(Age = list(dimension = "Age",
                                   unit = "yrs"),
                        Obs = list(dimension = "Clearance",
                                   unit = "dL/h/kg"),
                        Pred = list(dimension = "Clearance",
                                    unit = "dL/h/kg"),
                        Ratio = list(dimension = "Ratio",
                                     unit = "")
                        )
```
```{r show metadata, results='asis'}
knitr::kable(data.frame(Variable = c("Age", "Obs", "Pred", "Ratio"),
                        Dimension = c("Age", "Clearance", "Clearance", "Ratio"),
                        Unit = c("yrs", "dL/h/kg", "dL/h/kg", "")))
```

## 3.2. Minimal example
In the minimal example, only the basic `y` variable name is indicated. Here, `"Age"` was chosen for the boxplot.

```{r minimal example, fig.height=5, fig.width=7.5}
minMap <- BoxWhiskerDataMapping$new(y = "Age")

minBoxplot <- plotBoxWhisker(data = pkRatioData, 
                          metaData = pkRatioMetaData, 
                          dataMapping = minMap)
minBoxplot
```

## 3.3. Difference `x` vs `fill` input
As explained in section 2, `x` and/or `fill` can be provided.
If providing only `x`, the plot will use the `x` variable for aggregation and the boxplots will be disposed according to `x`. If providing `fill`, the plot will use the `fill` groupMapping for aggregation and the boxplots will be disposed around the same `x` but comparing the color filling. Consequently, the `fill` variable is useful when performing a double comparison.

In the example below, `"Country"` and `"Sex"` can both be used for comparison of `"Age"`.

```{r difference x vs fill}
xPopMap <- BoxWhiskerDataMapping$new(x = "Country",
                                     y = "Age")

xSexMap <- BoxWhiskerDataMapping$new(x = "Sex",
                                     y = "Age")

fillPopMap <- BoxWhiskerDataMapping$new(y = "Age",
                                        fill = "Country")

fillSexMap <- BoxWhiskerDataMapping$new(y = "Age",
                                        fill = "Sex")
xPopFillSexMap <- BoxWhiskerDataMapping$new(x = "Country",
                                            y = "Age",
                                            fill = "Sex")

xSexFillPopMap <- BoxWhiskerDataMapping$new(x = "Sex",
                                            y = "Age",
                                            fill = "Country")
```

```{r boxplot country x, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Country as x"}
plotBoxWhisker(data = pkRatioData,
               metaData = pkRatioMetaData,
               dataMapping = xPopMap)
```
```{r boxplot sex x, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Sex as x"}
plotBoxWhisker(data = pkRatioData,
               metaData = pkRatioMetaData,
               dataMapping = xSexMap)
```
```{r boxplot country fill, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Country as fill"}
plotBoxWhisker(data = pkRatioData,
               metaData = pkRatioMetaData,
               dataMapping = fillPopMap)
```
```{r boxplot sex fill, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Sex as fill"}
plotBoxWhisker(data = pkRatioData,
               metaData = pkRatioMetaData,
               dataMapping = fillSexMap)
```
```{r boxplot country x sex fill, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Country as x and Sex as fill"}
plotBoxWhisker(data = pkRatioData,
               metaData = pkRatioMetaData,
               dataMapping = xPopFillSexMap)
```
```{r boxplot country fill sex x, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Sex as x and Country as fill"}
plotBoxWhisker(data = pkRatioData,
               metaData = pkRatioMetaData,
               dataMapping = xSexFillPopMap)
```

## 3.4. Boxplot functions
In some cases, 5th and 95th percentiles are not wanted for whiskers for instance when a normal distribution is assumed and mean +/- 1.96 standard deviation would be preferred.
In these cases, it is easy to overwrite the default functions by specifying either using an home made function or directly using predifined functions as suggested in section 2.2.

In the 2 following examples, the boxplot will use the mean for the middle line and mean +/- 1.96 standard deviation for the whiskers:
```{r aggregation functions 1, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Country as x, Sex as fill and assuming normal distribution"}

normMap <- BoxWhiskerDataMapping$new(x = "Country",
                                     y = "Age",
                                     fill = "Sex",
                                     ymin = tlfStatFunctions$`mean-1.96sd`,
                                     middle = tlfStatFunctions$mean,
                                     ymax = tlfStatFunctions$`mean+1.96sd`)

normBoxplot <- plotBoxWhisker(data = pkRatioData,
                              metaData = pkRatioMetaData,
                              dataMapping = normMap)
normBoxplot
```

In this example, the boxplot use also mean +/- standard deviation for the box edges
```{r aggregation functions 2, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Country as x, Sex as fill and assuming normal distribution"}

normMap2 <- BoxWhiskerDataMapping$new(x = "Country",
                                     y = "Age",
                                     fill = "Sex",
                                     ymin = tlfStatFunctions$`mean-1.96sd`,
                                     lower = tlfStatFunctions$`mean-sd`,
                                     middle = tlfStatFunctions$mean,
                                     upper = tlfStatFunctions$`mean+sd`,
                                     ymax = tlfStatFunctions$`mean+1.96sd`)

normBoxplot2 <- plotBoxWhisker(data = pkRatioData,
                              metaData = pkRatioMetaData,
                              dataMapping = normMap2)
normBoxplot2
```

## 3.5. Outlyier functions

Default outliers are flagged when oustide the range from 25th percentiles - 1.5IQR to 75th percentiles + 1.5IQR as suggested by McGill and implemented by the current boxplot functions from ggplot (`geom_boxplot`).
However, these default can also be overwritten. 

In the following example, outliers will be flagged when values are out of the 10th-90th percentiles, while whiskers will go until these same percentiles:
```{r outlier function, fig.height=5, fig.width=7.5, fig.cap="Boxplot mapping Country as x, Sex as fill and assuming normal distribution"}

outlierMap <- BoxWhiskerDataMapping$new(x = "Country",
                                        y = "Age",
                                        fill = "Sex",
                                        ymin = tlfStatFunctions$`Percentile10%`,
                                        ymax = tlfStatFunctions$`Percentile90%`,
                                        minOutlierLimit = tlfStatFunctions$`Percentile10%`,
                                        maxOutlierLimit = tlfStatFunctions$`Percentile90%`)

outlierBoxplot <- plotBoxWhisker(data = pkRatioData,
                              metaData = pkRatioMetaData,
                              dataMapping = outlierMap)
outlierBoxplot
```

## 3.4. plotConfiguration of boxplots: `BoxWhiskerPlotConfiguration`

To define the properties of the boxes and points of the box whisker plots, a *BoxWhiskerPlotConfiguration* object can be defined to overwrite the default properties.
The *ribbons* and *points* fields will define how the boxes and outliers will be handled.

Using the previous example where country was defined in X and gender as a color.
```{r boxplot update configuration, fig.height=5, fig.width=7.5, fig.cap="Boxplot with updated plot configuration"}
# Define a PlotConfiguration object using smart mapping
boxplotConfiguration <- BoxWhiskerPlotConfiguration$new(data = pkRatioData,
                                                        metaData = pkRatioMetaData,
                                                        dataMapping = xPopFillSexMap)

# Change the properties of the box colors
boxplotConfiguration$ribbons$fill <- c("pink", "dodgerblue")
boxplotConfiguration$ribbons$color <- "black"

# Change the properties of the points (outliers)
boxplotConfiguration$points$size <- 2
boxplotConfiguration$points$shape <- Shapes$diamond

plotBoxWhisker(data = pkRatioData,
               metaData = pkRatioMetaData,
               dataMapping = xPopFillSexMap,
               plotConfiguration = boxplotConfiguration)
```

# 4. Derive use of `BoxWhiskerDataMapping`

Since the boxplot datamapping performs an aggregation of the data, it possible to get directly the resulting aggregated statistic as a table using `getBoxWhiskerLimits()`
Similarly, it can be used to flag any values out of a certain range using `getOutliers()`.

For instance, using the example from section 3.5., one can get the following results

```{r box plot measure, results='as.is'}
boxplotSummary <- outlierMap$getBoxWhiskerLimits(pkRatioData)

knitr::kable(boxplotSummary, digits = 2)
```

```{r get outliers, results='as.is'}
outliers <- outlierMap$getOutliers(pkRatioData)
outliers <- outliers[, c("Age", "minOutlierLimit", "maxOutlierLimit", "minOutliers", "maxOutliers")]

knitr::kable(utils::head(outliers), digits = 2)
```

