---
title: "PK Ratio Example"
author: "OSPSuiteR 2019"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pk-ratio-vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
load("../data/pkRatioDataExample.RData")
library(tlf)
```

# Introduction
The following vignette aims at showing an example workflow for performing PK Ratios plots and qualifications using the tlf-Library.
The very first step of the workflow is to load the `tlf` library and associated packages (so far `ggplot2`).
The 2 following steps are the definition of the `dataMapping` and `plotConfiguration`.

* `dataMapping` aims at linking which variable from the `data`, `metaData` should be plotted.
* `plotConfiguration` aims at defining how this map data should be plotted.

```{r}
useTheme(tlfTheme)
```

# Data
For this example, the `data` and its `metaData` will be called `pkRatioData` and `pkRatioMetaData`.
`pkRatioData` is expected to be a `data.frame` and `pkRatioMetaData` a `list` of lists containing, for each variable of `pkRatioData`, its unit and dimension.
Below are the first lines of the PK Ratio Dataset used in this example:

## Dataset
```{r, echo = FALSE, results='asis'}
# pkRatioData
pander::pandoc.table(head(pkRatioData))
```

# Data Mapping
To map which data to plot, `DataMapping` objects derived from `R6Class` can be used.
For PK Ratio plots, `PKRatioDataMapping` is the most suited mapping:
Default `x` is `"Age"` and default `y` is `"Ratio"`. However, this variables can be overwritten.
The class defines `pkRatioLines` as limits to be plotted in the PK Ratio plot with default values predefined.
Since `PKRatioDataMapping` inherits from an XY Group Mapping (`XYGDataMapping`), it allows as well the definition of groupings.

## Example of Definition of PK Ratio Data Mapping
```{r}
# Create a PK Ratio dataMapping
groups <- list(color = "Gender", shape = c("Dose", "Compound"))
dataMapping <- PKRatioDataMapping$new(x = "Age", y = "PKRatios", groupings = groups)


dataMapping$x
dataMapping$y
dataMapping$pkRatioLines
dataMapping$groupings$color
dataMapping$groupings$shape
```

# Plot Configuration
To configure the plot properties, `PlotConfiguration` objects derived from `R6Class` can be used.
They associate the following features:

* `LabelConfiguration`, inherited `R6Class` defining the labels of the plot (title, subtitles, x axis, y axis and watermark)
* `LegendConfiguration`, `R6Class` defining the legend properties of the plot
* `xAxisConfiguration`, `R6Class` defining the properties of the X axis (scale, limits, printed values)
* `yAxisConfiguration`, `R6Class` defining the properties of the Y axis (scale, limits, printed values)

For PK Ratio Plots, `PKRatioPlotConfiguration` is the most suited of `PlotConfiguration`
The most suited way is to provide `data`, `metaData` and `dataMapping` to get automatically the associated plot configuration. However, any of the properties can be overwritten (either by indicating the updated input or by redifing the value after the creation of the object).

## Example of Definition of PK Ratio Plot Configuration:

```{r}
# Create PK Ratio data Maping
dataMapping <- PKRatioDataMapping$new()

plotConfiguration <- PKRatioPlotConfiguration$new(
  data = pkRatioData,
  metaData = pkRatioMetaData,
  dataMapping = dataMapping
)

plotConfiguration
#plotConfiguration$legend
#plotConfiguration$xAxis
#plotConfiguration$yAxis
```

# Plot PK Ratios
The function `plotPKRatio` allows for an easy way of plotting PK Ratios with specific PK Ratio limits.

* Input arguments of `plotPKRatio` are the `data`, `metaData`, their `dataMapping` and  `plotConfiguration
* Output argument of `plotPKRatio` is the `plotObject` corresponding to a `ggplot` object.

If no `dataMapping` and `plotConfiguration` is provided, `plotPKRatio` uses a default mapping and configuration.

## Example of PK Ratio Plot workflow

```{r, fig.height=5, fig.width=7.5}
# Create PK Ratio data Mapping
pkRatioDataMapping <- PKRatioDataMapping$new(
  groupings = Groupings$new(color = "Gender",
                            shape = c("Dose", "Compound")))

# Create associated PK Ratio plot Configuration
pkRatioPlotConfiguration <- PKRatioPlotConfiguration$new(
  ylabel = "Simulated/Observed Cmax Ratio",
  data = pkRatioData,
  metaData = pkRatioMetaData,
  dataMapping = pkRatioDataMapping
)

# Define the printable ggplot object
plotObject <- plotPKRatio(
  data = pkRatioData, 
  metaData = pkRatioMetaData,
  dataMapping = pkRatioDataMapping, 
  plotConfiguration = pkRatioPlotConfiguration
)

plotObject

```

# Use of Themes to set standardized plots

Default values of plot configuration are set according to a `Theme` class defining the color and font of the every label, as well as the default properties of a plot (color palette and linetypes).
It is possible to create one's own `Theme` using `Theme$new()` and setting the properties as the default or the tlf environment using the funtion `useTheme()`.
Currently, 4 themes are defined in the library: `defaultTheme`, `tlfTheme`, `bwTheme` and `bigTheme`.

## The follwing example show the output of PK Ratio set from `defaultTheme`
```{r, fig.height=5, fig.width=7.5}
useTheme(defaultTheme)
# Create PK Ratio data Mapping
pkRatioDataMapping <- PKRatioDataMapping$new(
  groupings = Groupings$new(color = "Gender",
                            shape = c("Dose", "Compound")))

# Create associated PK Ratio plot Configuration
pkRatioPlotConfiguration <- PKRatioPlotConfiguration$new(
  data = pkRatioData,
  metaData = pkRatioMetaData,
  dataMapping = pkRatioDataMapping
)

# Define the printable ggplot object
plotObject <- plotPKRatio(
  data = pkRatioData, 
  metaData = pkRatioMetaData,
  dataMapping = pkRatioDataMapping, 
  plotConfiguration = pkRatioPlotConfiguration
)

plotObject

```

## The follwing example shows the output of PK Ratio set from balck and white `bwTheme`
```{r, fig.height=5, fig.width=7.5}
useTheme(bwTheme)
# Create PK Ratio data Mapping
pkRatioDataMapping <- PKRatioDataMapping$new(
  groupings = Groupings$new(color = "Gender",
                            shape = c("Dose", "Compound")))

# Create associated PK Ratio plot Configuration
pkRatioPlotConfiguration <- PKRatioPlotConfiguration$new(
  data = pkRatioData,
  metaData = pkRatioMetaData,
  dataMapping = pkRatioDataMapping
)

# Define the printable ggplot object
plotObject <- plotPKRatio(
  data = pkRatioData, 
  metaData = pkRatioMetaData,
  dataMapping = pkRatioDataMapping, 
  plotConfiguration = pkRatioPlotConfiguration
)

plotObject

```

# Qualification of PK Ratios
The qualification of the PK Ratios can be performed using `getPKRatioMeasure`.
This function return a `data.frame` with the PK ratios within specific ranges.
As a default, these ranges are within 1.5 and 2 folds. However, they can be updated using the option `ratioLimits =` when running the function.

```{r, results='asis'}
# Test of getPKRatioMeasure
PKRatioMeasure <- getPKRatioMeasure(data = pkRatioData, 
                                    dataMapping = pkRatioDataMapping)

pander::pandoc.table(t = PKRatioMeasure,
                     caption = "Qualification of PK Ratios")
```
