---
title: "PK Ratio Example"
author: "OSPSuiteR 2019"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pk-ratio-vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tlf)
```

# 1. Introduction
## 1.1. Objectives
The following vignette aims at documenting and illustrating workflows for producing PK ratio plots using the `tlf`-Library.

## 1.2. Prerequisite
This vignette focuses PK ratio plots examples. Detailed documentation on typical `tlf` workflow, use of `AgregationSummary`, `DataMapping`, `PlotConfiguration` and `Theme` can be found in `vignette("tlf-workflow")`.

# 2. Illustration of basic PK ratio plots
## 2.1. Data
To illustrate the workflow leading to performing PK ratio plots. The `pkRatioDataExample.RData` example can be loaded from the `data` folder. 
It includes the dataset `pkRatioData` and its metaData `pkRatioMetaData`.

```{r, results='asis'}
# Load example
load("../data/pkRatioDataExample.RData")

# pkRatioData
knitr::kable(utils::head(pkRatioData), digits = 2)

# pkRatioMetaData is a list of variables contianing a lists with unit, dimension and lloq for each
# Unit and dimension of pkRatioData are consequently the following:
knitr::kable(data.frame(unit = t(rbind(sapply(pkRatioMetaData, function(x){x$unit}))),
                        dimension = t(rbind(sapply(pkRatioMetaData, function(x){x$dimension})))))
```


## 2.2. `plotPKRatio`
The function plotting PK ratios is: `plotPKRatio`. 
Basic documentation of the function can be found using: `?plotPKRatio`.
The typical usage of this function is: `plotPKRatio(data, metaData = NULL, dataMapping = NULL, plotConfiguration = NULL)`.
The output of the function is a `ggplot` object.
It can be seen from this usage that only `data` is a necessary input.
Default set ups are used for `metaData`, `dataMapping` and `plotConfiguration` within the call of `plotPKRatio`.
A `metaDataHelper` function is first used if `metaData` is not provided. The function creates a `metaData` structure with empty values, then identifies the variable names that are commonly used to linik them to their usual `unit` and `dimension`. 
For instance, `Age` can be identified and its most common unit `yrs` can be associated to the plot.
`PKRatioDataMapping` is initialized if not provided, defining a standard mapping for plotting `"Age"` as `x` variable and `"Ratio"` as `y` variable.
Then, `PKRatioPlotConfiguration` is initialized if not provided, defining a standard configuration with `PK Ratio Plot` as title, the current date as subtitle and using predifned fonts as defined by the current theme.

## 2.3. Minimal example
The minimal example can work using directly the function `plotPKRatio(pkRatioData)`.
The output is exactly the same as calling `plotPKRatio(data = pkRatioData, metaData = NULL, dataMapping = NULL, plotConfiguration = NULL)`.

```{r, fig.height=5, fig.width=7.5}
plotPKRatio(data = pkRatioData)
```

## 2.4. Examples with `dataMapping`
For PK ratio, the `dataMapping` class `PKRatioDataMapping` includes 4 fields: `x`, `y`, `groupMapping` and `pkRatioLines`.
`x` and `y` define which variables from the data will be plotted in X- and Y-axes, `groupMapping` is a class mapping which aesthtic property will split which variables, and `pkRatioLines` defines horizontal lines performed in PK ratio plots.

### 2.4.1 `groupMapping`
The `groupMapping` input is assumed a `GroupMapping` class, meaning it should be initialized prior the dataMapping.
To simplify this process, `groupMapping` input can directly be initilized within the `dataMapping`.
Consequently, The following examples are identical:
```{r}
# Two-step process
colorMapping <- GroupMapping$new(color = "Gender")
dataMappingA <- PKRatioDataMapping$new(x = "Age", 
                                      y = "Ratio", 
                                      groupMapping = colorMapping)

print(dataMappingA$groupMapping$color$label)

# One-step process
dataMappingB <- PKRatioDataMapping$new(x = "Age", 
                                      y = "Ratio", 
                                      color = "Gender")

print(dataMappingB$groupMapping$color$label)

```

Then, in this example, `plotPKRatio` can use the groupMapping to split the data by "Gender" and associate different colors to each "Gender":
```{r, fig.height=5, fig.width=7.5}
plotPKRatio(data = pkRatioData, 
            dataMapping = dataMappingB)
```

Multiple groupMappings can be performed for PK ratio: data can be regrouped by `color`, `shape` and/or `size`.
The next example uses 2 groups in the groupMapping:
One group splits "Gender" by `color`, the other splits `shape` by "Amount" and "Compound".
```{r, fig.height=5, fig.width=7.5}
dataMapping2groups <- PKRatioDataMapping$new(x = "Age",
                                             y = "Ratio",
                                             color = "Gender",
                                             shape = c("Dose", "Compound"))
plotPKRatio(data = pkRatioData, 
            dataMapping = dataMapping2groups)
```

The last examples uses another feature available in the `groupMapping` class. The class can be initilized using a `data.frame` where the last column of the data.frame will be used to split the data. In the following example, the data.frame is the following:

```{r, echo=FALSE, results='asis'}
groupDataFrame <- data.frame(Dose = c(50, 100, 50, 100, 50, 100, 50, 100), 
                             Compound = c("Aspirin", "Aspirin", "Vancomycin", "Vancomycin", "Aspirin", "Aspirin", "Vancomycin", "Vancomycin"),
                             Gender = c("M", "M", "M", "M", "F", "F", "F", "F"),
                             Group = c("Males infused with 50mg of Aspirin", "Males infused with 100mg of Aspirin", 
                                       "Males infused with 50mg of Vancomycin", "Males infused with 100mg of Vancomycin",
                                       "Females infused with 50mg of Aspirin",  "Females infused with 100mg of Aspirin", 
                                       "Females infused with 50mg of Vancomycin", "Females infused with 100mg of Vancomycin"))
knitr::kable(groupDataFrame)
```

The `dataMapping` introduced below will split the `color` and `shape` using the data frame.
```{r, fig.height=5, fig.width=7.5}
dataMappingDF <- PKRatioDataMapping$new(x = "Age",
                                             y = "Ratio",
                                             color = groupDataFrame,
                                             shape = groupDataFrame)
plotPKRatio(data = pkRatioData, 
            dataMapping = dataMappingDF)
```


### 2.4.2 `pkRatioLines`
In PK ratio examples, usually horizontal lines are added allowing to flag values in and out of the [0.67-1.5] as well as [0.5-2.0] ranges.
The value mapping these horizontal lines was predifined as the vector `c(1.0, )`.
Consequently, for any default `PKRatioDataMapping`, you have:
```{r}
linesMapping <- PKRatioDataMapping$new()
linesMapping$pkRatiolines
```

Overwriting these value is possible by updating the value either when initializing the mapping or afterwards.
For instance:
```{r, fig.height=5, fig.width=7.5}
linesMapping <- PKRatioDataMapping$new(pkRatioLines = c(1, 2, 3),
                                       x = "Age", 
                                       y = "Ratio",
                                       color = "Gender")
plotPKRatio(data = pkRatioData, 
            dataMapping = linesMapping)

```

## 2.5. Plot Configuration

To configure the plot properties, `PlotConfiguration` objects derived from `R6Class` can be used.
They associate the following features:

* `LabelConfiguration`, inherited `R6Class` defining the labels of the plot (title, subtitles, x axis, y axis)
* `LegendConfiguration`, `R6Class` defining the legend properties of the plot
* `BackgroundConfiguration`, `R6Class` defining the background properties of the plot (color, grid and watermark)
* `xAxisConfiguration`, `R6Class` defining the properties of the X axis (scale, limits, printed values)
* `yAxisConfiguration`, `R6Class` defining the properties of the Y axis (scale, limits, printed values)

Plot configuration uses default properties from theme but can be overwritten (either by indicating the updated input or by redifing the value after the creation of the object).

## 2.5.1 Themes

Plot configuration uses default properties from an `R6 class` Theme. 
Currently, 3 main themes are predefined and available. However, it is possible to create one's own theme by initiliazing the `Theme` class.

To set theme, use function `useTheme` as introduced in the following example:
```{r, fig.height=5, fig.width=7.5}
useTheme(bwTheme)
plotPKRatio(data = pkRatioData)
```
```{r, fig.height=5, fig.width=7.5}
useTheme(tlfTheme)
plotPKRatio(data = pkRatioData)
```


## 2.5.2 Labels
Default labels are implemented for titles and axes. 
The default title is the type of the plot here "PK Ratio Plot", the default subtitle is the current date, as for the axes metaData are searched and the `dimension [unit]` of the metaData is provided as default.
However, it is possible to overwrite the text using the follwong example:
```{r, fig.height=5, fig.width=7.5}
labelConfiguration <- PKRatioPlotConfiguration$new(title = "New Title",
                                                   subtitle = "New subtitle",
                                                   xlabel = "New X-axis",
                                                   ylabel = "New Y-axis")
plotPKRatio(data = pkRatioData,
            plotConfiguration = labelConfiguration)
```

If needed, it is also possible to update the font properties either by using a `Label` class as input or by updating the `Label` output of the plotConfiguration.
For instance, the 2 following examples are equivalent:
```{r, fig.height=5, fig.width=7.5}
labelConfiguration1 <- PKRatioPlotConfiguration$new(title = Label$new(text = "Title as Label", 
                                                                     font = Font$new(color = "red")))
labelConfiguration2 <- PKRatioPlotConfiguration$new(title = "Title as Label")
labelConfiguration2$title$font$color <- "red"

isTRUE(labelConfiguration1$title$font$color == labelConfiguration2$title$font$color)
```

## 2.5.3 Axes

## 2.5.4 Background

## 2.5.2 Legends

# 3. Qualification of PK Ratios
The qualification of the PK Ratios can be performed using `getPKRatioMeasure`.
This function return a `data.frame` with the PK ratios within specific ranges.
As a default, these ranges are within 1.5 and 2 folds. However, they can be updated using the option `ratioLimits =` when running the function.

```{r, results='asis'}
# Test of getPKRatioMeasure
PKRatioMeasure <- getPKRatioMeasure(data = pkRatioData)

knitr::kable(x = PKRatioMeasure,
             caption = "Qualification of PK Ratios")
```

# 4. Going further: DDI Ratio plot examples
PK ratio plots and DDI ratio plots are very similar in their structure. 
The difference is the plotting of diagonal lines and limits as defined by the Guest et al. equation.
These plots introduce 4 new fields in the method: `xmin`, `xmax`, `ddiRatioLines`, `deltaGuest`.
`xmin` and `xmax` defining the range where the diagonal lines and Guest limits should be plotted, the default range is [0.1-10].
`ddiRatioLines` the values of the ratios (similar as `pkRatioLines` the dfault is `c(1,2,0.5)`). 
`deltaGuest` is the value of `delta` used in the Guest equation (default set as 1).

The definition of such lines and limits is pre-defined within the dataMapping class for DDI ratios: `DDIRatioDataMapping`, as shown below:
```{r}
linesMapping <- DDIRatioDataMapping$new()
```

Then, keeping the workflow as provided in the PK ratio examples, DDI ratio plot can be performed in very few steps:
```{r, fig.height=5, fig.width=7.5}
useTheme(tlfTheme)

# Process data
pkRatioData$Observed <- pkRatioData$Observed/10
pkRatioData$Simulated <- pkRatioData$Simulated/10

# Define Mapping
ddiMapping <- DDIRatioDataMapping$new(x = "Observed",
                                      y = "Simulated",
                                      color = "Gender",
                                      shape = c("Dose", "Compound"))
# Define configuration
xAxis <- XAxisConfiguration$new(scale = Scaling$log10)
yAxis <- YAxisConfiguration$new(scale = Scaling$log10)
ddiConfiguration <- DDIRatioPlotConfiguration$new(xlabel = "Obs",
                                                  ylabel = "Pred",
                                                  data = pkRatioData,
                                                  metaData = pkRatioMetaData,
                                                  dataMapping = ddiMapping,
                                                  xAxis = xAxis,
                                                  yAxis = yAxis)
# Overwrite default size of points
ddiConfiguration$legend$values$size <- 2

# Plot
plotDDIRatio(data = pkRatioData,
             metaData = pkRatioMetaData,
             dataMapping = ddiMapping,
             plotConfiguration = ddiConfiguration)
```