---
title: "Plot configuration"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{plot-configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tlf)
```


This vignette tackles about *PlotConfiguration* objects and their implementation within the `tlf`-library.

# Introduction

*PlotConfiguration* objects are R6 class objects that define plot properties. 

To create such an object, the method *new()* is required. 
Multiple arguments can be passed on the method in order to directly define the properties. 
The properties default values can be handled and managed using the concept of themes.

The class possesses also a *print()* method that will print the configuration of the properties below.

As the example below illustrates, the following properties are accounted for by the *PlotConfiguration* class. 
Each of these properties is managed by a different R6 class object detailed in the next sections:

- *labels*
- *background*
- *xAxis*/*yAxis*
- *legend*
- *save*

```{r}
myConfiguration <- PlotConfiguration$new()
# myConfiguration$print()

myConfiguration <- PlotConfiguration$new(title = "my title",
                                         watermark = "my watermark")
# myConfiguration$print()
```

When a `ggplot` object is initialized using the `tlf` method *initializePlot*, a *plotConfiguration* field of is implanted to the plot object.
The *plotConfiguration* field is class or subclass of *PlotConfiguration* that aims at simplifying and clarifying the plot properties mentioned above.

*PlotConfiguration* objects can also be pass on the method *initializePlot* as shown below:

```{r}
emptyPlot <- initializePlot()

myEmptyPlot <- initializePlot(myConfiguration)
```
```{r, echo=FALSE, fig.cap="left: emptyPlot; right: myEmptyPlot", fig.width=7.5}
gridExtra::grid.arrange(emptyPlot, myEmptyPlot, ncol=2)
```

# Label configuration: *labels*

## Label and Label configuration objects

The field *labels* from the *PlotConfiguration* object is a *LabelConfiguration* R6 class object.
It defines the label properties of the following plot captions:

- title
- subtitle
- xlabel
- ylabel

Each field is a *Label* R6 class object that associate a text with font properties:
```{r}
myEmptyLabel <- Label$new()
myEmptyLabel$print()

myLabel <- Label$new(text = "some text",
                     color = "blue")
myLabel$print()
```

When initializating a *LabelConfiguration* or *PlotConfiguration* object, *Label* or/and *character* objects can be passed on.
*Character* objects will be converted into *Label* objects internally using the current theme.

```{r}
myRedPlotLabel <- LabelConfiguration$new(title = Label$new(text = "my title", 
                                                           color = "red"))
myRedPlotLabel$title$print()
myRedPlotLabel$print()

myPlotLabel <- LabelConfiguration$new(title = "my title")
myPlotLabel$title$print()
myPlotLabel$print()
```

It can be noted that in the latter case, the label configuration is equivalent to the label configuration defined in the previous plot configuration "myConfiguration":

```{r}
myConfiguration$labels$title$print()
myConfiguration$labels$print()
```

## Label configuration in plots

The effect of label configuration in plots is straightforward:
```{r}
plotConfigurationLabel1 <- PlotConfiguration$new(title = "Title",
                                                     subtitle = "Subtitle",
                                                     xlabel = "x label",
                                                     ylabel = "y label")

pLab1 <- initializePlot(plotConfigurationLabel1)

plotConfigurationLabel2 <- PlotConfiguration$new(title = Label$new(text = "Title", size = 12, color = "deepskyblue4"),
                                                 subtitle = Label$new(text = "Subtitle", size = 11, color = "steelblue"),
                                                 xlabel = Label$new(text = "x label", size = 10, color = "dodgerblue2"),
                                                 ylabel = Label$new(text = "y label", size = 10, color = "dodgerblue2"))

pLab2 <- initializePlot(plotConfigurationLabel2)
```
```{r, echo=FALSE, fig.cap="left: pLab1; right: pLab2", fig.width=7.5}
gridExtra::grid.arrange(pLab1, pLab2, ncol=2)
```

## Changing plot labels

After creating a plot, it is possible to change its label configuration using the `tlf` method *setPlotLabels*.
The metod requires the plot object and which label property to update.
Similar to the construction of the Label configuration, *Label* or/and *character* objects can be passed on.

Using the previous example, *pLab2*, 
```{r}
pLab2$plotConfiguration$labels$title$print()

pLab2Title <-  setPlotLabels(pLab2, title = "new title")
pLab2RedTitle <-  setPlotLabels(pLab2, title = Label$new(text = "new title", color = "red"))

pLab2Title$plotConfiguration$labels$title$print()
pLab2RedTitle$plotConfiguration$labels$title$print()
```
```{r, echo=FALSE, fig.cap="left: pLab2Title; right: pLab2RedTitle", fig.width=7.5}
gridExtra::grid.arrange(pLab2Title, pLab2RedTitle, ncol=2)
```

## Smart plot configurations

Smart plot labels are available for initializing *PlotConfiguration* objects. 
The principle is to provide in advance the *data* that will be used within the plot.
As a consequence, three optional arguments can be passed on *PlotConfiguration* initialization: *data*, *metaData* and *dataMapping*.
If no label is specifically defined, the smart configuration will fetch x and y labels within *data* and *metaData* names based on the *dataMapping*.
*dataMapping* uses also smart functions that will check the input data.frame if not specifically initialized and will use variables named "x" and "y" as *x* and *y*.

The 4 examples below illustrates the smart configurations. *pSmart1* and *pSmart2* will turn out the same, while *pSmart3* will use the information from *metaData* to write the x and y labels. *pSmart4* will overwrite the y label and title based on the input.
```{r}
time <- seq(0,20, 0.1)
myData <- data.frame(x = time,
                     y = 2*cos(time))
myMetaData <- list(x = list(dimension = "Time",
                            unit = "min"),
                   y = list(dimension = "Amplitude",
                            unit = "cm"))
myMapping <- XYGDataMapping$new(x = "x",
                                y = "y")

smartConfig1 <- PlotConfiguration$new(data = myData)
smartConfig2 <- PlotConfiguration$new(data = myData,
                                      dataMapping = myMapping)
smartConfig3 <- PlotConfiguration$new(data = myData,
                                      metaData = myMetaData)
smartConfig4 <- PlotConfiguration$new(title = Label$new(text = "Cosinus", size = 14),
                                      ylabel = "Variations",
                                      data = myData,
                                      metaData = myMetaData)

pSmart1 <- initializePlot(smartConfig1) 
pSmart2 <- initializePlot(smartConfig2) 
pSmart3 <- initializePlot(smartConfig3) 
pSmart4 <- initializePlot(smartConfig4) 
```

```{r, echo=FALSE, fig.cap="left: pSmart1; right: pSmart2", fig.width=7.5}
gridExtra::grid.arrange(pSmart1, pSmart2, ncol=2)
```

```{r, echo=FALSE, fig.cap="left: pSmart3; right: pSmart4", fig.width=7.5}
gridExtra::grid.arrange(pSmart3, pSmart4, ncol=2)
```

Since all of the `tlf` plots are using internally *initializePlot* if a previous plot is not provided, the smart configurations can directly be used through the plot functions. Consequently, if *scatter1* will provide simple x and y labels, *scatter2* will name these labels after the metaData properties.
As for *scatter3* and *scatter4*, they will use the plotConfiguration defined by *smartConfig4* and lead to the exact same plot.
```{r}
scatter1 <-  addScatter(data = myData)
scatter2 <-  addScatter(data = myData,
                        metaData = myMetaData)

scatter3 <-  addScatter(data = myData,
                        plotConfiguration = smartConfig4)
scatter4 <- initializePlot(smartConfig4)
scatter4 <- addScatter(data = myData,
                       plotObject = scatter4)
```

```{r, echo=FALSE, fig.cap="left: scatter1; right: scatter2", fig.width=7.5}
gridExtra::grid.arrange(scatter1, scatter2, ncol=2)
```

```{r, echo=FALSE, fig.cap="left: scatter3; right: scatter4", fig.width=7.5}
gridExtra::grid.arrange(scatter3, scatter4, ncol=2)
```

# Background configuration: *background*

The field *background* from the *PlotConfiguration* object is a *BackgroundConfiguration* R6 class object.
It defines the following plot properties:

- inner and outer backgrounds
- grid
- watermark

Inner, outer backgrounds and grid are all *BackgroundElementConfiguration* R6 class object that associate one or multiple aesthetic properties to the background element.
Watermark is a *Label* R6 class object to which it is also possible to set the angle and transparency.

When initializating a *BackgroundConfiguration* or *PlotConfiguration* object, the background properties and watermark can be passed on.

```{r}
myEmptyBackgroundConfiguration <- BackgroundConfiguration$new()
myEmptyBackgroundConfiguration$print()
```

## Background configuration in plots

The effect of background configuration in plots is quite straightforward:
```{r}
plotConfigurationBackground1 <- PlotConfiguration$new(watermark = "My Watermark")

pBack1 <- initializePlot(plotConfigurationBackground1)

plotConfigurationBackground2 <- PlotConfiguration$new(
  watermark = Label$new(text = "Hello world", color = "goldenrod4", size = 8),
  background = BackgroundConfiguration$new(outerBackground = BackgroundElementConfiguration$new(fill = "lemonchiffon"),
                                           innerBackground = BackgroundElementConfiguration$new(fill = "lemonchiffon", 
                                                                                                color = "goldenrod3"),
                                           grid = BackgroundElementConfiguration$new(color = "goldenrod3")))

pBack2 <- initializePlot(plotConfigurationBackground2)
```

```{r, echo=FALSE, fig.cap="left: pBack1; right: pBack2", fig.width=7.5}
gridExtra::grid.arrange(pBack1, pBack2, ncol=2)
```

## Changing background properties

After creating a plot, it is possible to change its background configuration using the following `tlf` methods:

- *setBackground*
- *setGrid*
- *setWatermark*

### Set Background

The metod *setBackground* requires the plot object and which inner/outer background properties to update.

Using the previous example, *scatter1*, 
```{r}
scatter1$plotConfiguration$background$print()

scatterNewBackground1 <-  setBackground(scatter1, 
                             outerBackgroundFill = "darkolivegreen1",
                             fill = "darkolivegreen2",
                             color = "darkgreen",
                             linetype = "blank")
scatterNewBackground2 <-  setBackground(scatter1, 
                             outerBackgroundFill = "lemonchiffon",
                             fill = "lemonchiffon",
                             size = 2,
                             linetype = "solid")

scatterNewBackground1$plotConfiguration$background$print()
scatterNewBackground2$plotConfiguration$background$print()
```

```{r, echo=FALSE, fig.cap="top: scatter1; bottom left: scatterNewBackground1; bottom right: scatterNewBackground2", fig.width=7.5}
gridExtra::grid.arrange(grobs = list(scatter1, scatterNewBackground1, scatterNewBackground2), layout_matrix=rbind(c(1,1), c(2,3)))
```

### Set Grid

The metod *setGrid* is very similar and requires the plot object and which grid properties to update.

Using the previous examples, *scatterNewBackground1* and *scatterNewBackground2*:
```{r}
scatterNewBackground1$plotConfiguration$background$grid$print()

# Remove grid
scatterNewGrid1 <-  setGrid(scatterNewBackground1, 
                             linetype = "blank")

scatterNewGrid1$plotConfiguration$background$grid$print()

scatterNewGrid2 <-  setGrid(scatterNewBackground2, 
                             color = "darkgreen",
                             size = 0.5,
                             linetype = "dotted")

scatterNewGrid2$plotConfiguration$background$grid$print()
```

```{r, echo=FALSE, fig.cap="left: scatterNewBackground1; right: scatterNewGrid1", fig.width=7.5}
gridExtra::grid.arrange(scatterNewBackground1, scatterNewGrid1, ncol=2)
```

```{r, echo=FALSE, fig.cap="left: scatterNewBackground2; right: scatterNewGrid2", fig.width=7.5}
gridExtra::grid.arrange(scatterNewBackground2, scatterNewGrid2, ncol=2)
```


### Set Watermark

The metod *setWatermark* requires the plot object, the watermark and its properties to update.
In the properties, *alpha* is a value between 0 and 1 managing transparency of the watermark.
The closer to 0, the more transparent the watermark will be. 
The closer to 1, the more opaque the watermark will be.
The propert *angle*, is the angle of the watermark in degrees.

Using the previous examples, *pBack2* and *scatterNewGrid2*, 
```{r}
pBack2$plotConfiguration$background$watermark$print()

pWatermark1 <-  setWatermark(pBack2, 
                            watermark = "Hi !!",
                            alpha = 1,
                            angle = 0)

pWatermark2 <-  setWatermark(pBack2,
                            alpha = 1,
                            angle = 45)

pWatermark1$plotConfiguration$background$watermark$print()
pWatermark2$plotConfiguration$background$watermark$print()
```

```{r}
scatterNewGrid2$plotConfiguration$background$watermark$print()

scatterWatermark <-  setWatermark(scatterNewGrid2, 
                                  watermark = Label$new(text = "Confidential", color = "firebrick"))

scatterWatermark$plotConfiguration$background$watermark$print()
```

```{r, echo=FALSE, fig.cap="top: pBack2; bottom left: pWatermark2; bottom right: pWatermark2", fig.width=7.5}
gridExtra::grid.arrange(grobs = list(pBack2, pWatermark1, pWatermark2), layout_matrix = rbind(c(1,1),c(2,3)))
```

```{r, echo=FALSE, fig.cap="left: scatterNewGrid2; right: scatterWatermark", fig.width=7.5}
gridExtra::grid.arrange(scatterNewGrid2, scatterWatermark, ncol=2)
```

# X/Y axes configurations: *xAxis*/*yAxis*

The fields *xAxis* and *yAxis* from the *PlotConfiguration* object are a *XAxisConfiguration* and *YAxisConfiguration* R6 class objects.
They define the following plot properties:

- scale
- limits
- ticks
- ticklabels

The property *scale* is a character string corresponding the axis scale.
Available scales are `r paste0(c('"', paste0(Scaling, collapse = '", "'), '"'), collapse="")` and can be accessed using the enum *Scaling*.
The property *limits* is a vector defining the range of the axis. 
The value *"default"* will leave the management of the limits to R.
The properties *ticks* and *ticklabels* are vectors defining the plot ticks and their labels. These vectors are required to have the same length.
The value *"default"* will leave the management of the limits to R.

When initializating a *XAxisConfiguration*, *YAxisConfiguration* or *PlotConfiguration* objects, the axis properties can be passed on.


```{r}
myEmptyAxisConfiguration <- XAxisConfiguration$new()
myEmptyAxisConfiguration$print()

myAxisConfiguration <- XAxisConfiguration$new(scale = Scaling$log10,
                                              limits = c(1, 100))
myAxisConfiguration$print()
```

## Axis configuration in plots

## Changing axis properties

The metods *setXAxis* and *setYAxis* requires the plot object, and the axis properties to update.

Using the previous example, *scatter1*: 
```{r}
scatter1$plotConfiguration$xAxis$print()
scatter1$plotConfiguration$yAxis$print()

scatterXlog <-  setXAxis(scatter1,
                          scale = Scaling$log10)

scatterXlog$plotConfiguration$xAxis$print()

scatterXTicks <-  setXAxis(scatter1,
                           limits = c(0, 6*pi),
                           ticks = seq(0, 6*pi, pi),
                           ticklabels = parse(text = c("0", "pi", paste0(seq(2, 6), "*pi"))))

scatterXTicks$plotConfiguration$xAxis$print()

```

```{r, echo=FALSE, fig.cap="top: scatter1; bottom left: scatterXlog; bottom right: scatterXTicks", fig.width=7.5}
gridExtra::grid.arrange(grobs = list(scatter1, scatterXlog, scatterXTicks), layout_matrix = rbind(c(1,1),c(2,3)))
```

# Legend configuration: *legend*

The field *legend* from the *PlotConfiguration* object is a *LegendConfiguration* R6 class object.
It defines the following plot properties:

- title
- position
- caption

The property *title* is a character string corresponding the legend title.
The property *position* is a character string corresponding the legend position.
Available legend positions are `r paste0(c('"', paste0(LegendPositions, collapse = '", "'), '"'), collapse="")` and can be accessed using the enum *LegendPositions*.
The property *caption* is a data.frame defining the caption properties of the legend. 

When initializating a *LegendConfiguration* or *PlotConfiguration* object, the background properties and watermark can be passed on.

```{r}
myEmptyLegendConfiguration <- LegendConfiguration$new()
myEmptyLegendConfiguration$print()

myLegendConfiguration <- LegendConfiguration$new(title = "Legend Title",
                                                 position = LegendPositions$insideTopRight)
myLegendConfiguration$print()
```

## Legend configuration in plots

The effect of legend configuration in plots is straightforward.
Using the previous example with *myData*, initializing the plot won't print the legend on the plot since no data are plotted, but will keep in memory the configuration when adding layers.
The legend caption is updated every type a new layer is added. 
The atom plots also allow to manually indicates desired legend properties for the corresponding layer.

```{r}
plotConfigurationLegend1 <- PlotConfiguration$new()
plotConfigurationLegend1$legend$print()

emptyLegend1 <- initializePlot(plotConfigurationLegend1)
emptyLegend1$plotConfiguration$legend$print()

scatterLegend1 <- addScatter(data = myData,
                             plotConfiguration = plotConfigurationLegend1)
scatterLegend1$plotConfiguration$legend$print()

scatterLegend2 <- addLine(y = c(-1, 0, 1),
                          plotObject = scatterLegend1)
scatterLegend2$plotConfiguration$legend$print()

scatterLegend3 <- addScatter(data = myData,
                             caption = "Cosinus data",
                             color = "darkblue",
                             size = 2,
                             plotConfiguration = plotConfigurationLegend1)
scatterLegend3 <- addLine(y = c(-1, 0, 1),
                          caption = "Some lines",
                          color = "coral3",
                          linetype = "solid",
                          plotObject = scatterLegend3)
scatterLegend3$plotConfiguration$legend$print()
```

```{r, echo=FALSE, fig.cap="top left: emptyLegend1; top right: scatterLegend1; bottom left: scatterLegend2; bottom right: scatterLegend3", fig.width=7.5}
gridExtra::grid.arrange(emptyLegend1, scatterLegend1, scatterLegend2, scatterLegend3, ncol=2)
```

```{r}
plotConfigurationLegend1 <- PlotConfiguration$new(legendTitle = "my legend",
                                                  legendPosition = LegendPositions$outsideTopLeft)
plotConfigurationLegend1$legend$print()

emptyLegend1 <- initializePlot(plotConfigurationLegend1)
emptyLegend1$plotConfiguration$legend$print()

scatterLegend1 <- addScatter(data = myData,
                             plotConfiguration = plotConfigurationLegend1)
scatterLegend1$plotConfiguration$legend$print()

scatterLegend2 <- addLine(y = c(-1, 0, 1),
                          plotObject = scatterLegend1)
scatterLegend2$plotConfiguration$legend$print()

scatterLegend3 <- addScatter(data = myData,
                             caption = "Cosinus data",
                             color = "darkblue",
                             size = 2,
                             plotConfiguration = plotConfigurationLegend1)
scatterLegend3 <- addLine(y = c(-1, 0, 1),
                          caption = "Some lines",
                          color = "coral3",
                          linetype = "solid",
                          plotObject = scatterLegend3)
scatterLegend3$plotConfiguration$legend$print()
```

```{r, echo=FALSE, fig.cap="top left: emptyLegend1; top right: scatterLegend1; bottom left: scatterLegend2; bottom right: scatterLegend3", fig.width=7.5}
gridExtra::grid.arrange(emptyLegend1, scatterLegend1, scatterLegend2, scatterLegend3, ncol=2)
```

## Changing plot legend

### Changing a legend title

After creating a plot, it is possible to change its legend title using the `tlf` method *setLegendTitle*.
The metod requires the plot object and the text of title to update with.

Using *scatterLegend3* from the previous example:
```{r}
scatterLegend3$legend$title

scatterLegend4<- setLegendTitle(scatterLegend3, "My new legend")
scatterLegend4$legend$title
```

```{r, echo=FALSE, fig.cap="left: scatterLegend3; right: scatterLegend4", fig.width=7.5}
gridExtra::grid.arrange(scatterLegend3, scatterLegend4, ncol=2)
```

### Changing a legend position

After creating a plot, it is also possible to change its legend position using the `tlf` method *setLegendPosition*.
The metod requires the plot object and the new position of the legend to update with.

Using *scatterLegend3* from the previous example:
```{r}
scatterLegend3$legend$position

scatterLegend3Inside<- setLegendPosition(scatterLegend3, LegendPositions$insideTopRight)
scatterLegend3Inside$legend$position
```

```{r, echo=FALSE, fig.cap="left: scatterLegend3; right: scatterLegend3Inside", fig.width=7.5}
gridExtra::grid.arrange(scatterLegend3, scatterLegend3Inside, ncol=2)
```
  
### Changing a legend caption

# Usage of *theme*

# *PlotConfiguration* derived classes

